Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-05-10 writing/manuscript.Rmd
updated abstract to tws draft abstract; added refs

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on fine-scale movement and habitat selection. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature on mammals' movement rates and habitat selection. We then leverage the estimated effects to predict how and where mammals will move in British Columbia (BC), Canada, throughout the 21^st^ century. We show that changes in temperature altered movement phenology, and that climate change will impact mammals' movement rates and habitat selection throughout the century, but that there are no common trends across all study species. Consequently, we suggest that species' responses be interpreted carefully and considered jointly when extrapolating the community-level consequences of climate change."
```

@@

\noindent \textbf{References}: 140 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021]. Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

Our ability to respond to current and future changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014]. Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021]. Similarly, protecting a third of the world's ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021]. The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. The width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas. Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@duncan_predicting_2012].

>>>>

```{r, echo=FALSE}
abstract <- "Recent widespread warming has caused many terrestrial mammals to change when, how, and where they move, with cascading effects on individuals' fitness and habitat selection at the population, community, and ecosystem levels. Previous research has quantified the effects of temperature on mammalian movement behaviour, but few have quantified the effects of future climate change on mammalian movement. We address this gap by using Continuous-Time Movement Models and Hierarchical Generalized Additive Models to quantify the effects of temperature the probability of moving, the speed when moving, and the habitat selection of six species throughout British Columbia, Canada. We show that temperature is a strong determinant of when, how, and where mammals move, and that climate change will impact mammals' movement rates and habitat selection throughout the 21st century. We found no common effects of temperature on the species’ movement rates, so we suggest that estimated responses be considered jointly when making inferences about how climate change will impact ecological communities. However, habitat selection strength decreased for all species within their current ranges, suggesting that all six species will likely shift their range in the coming decades. Our findings are particularly concerning as increasingly common “climate weirding” pushes species to atypical ranges and behaviors. We conclude by placing our findings within a holistic and collaborative framework while providing suggestions for conservation and future research."
```

@@

\noindent \textbf{References}: 142 (updated on 2025-05-09) <!-- counted by using movement ecology csl -->

@@

\noindent We have demonstrated that temperature is an important determinant of whether, how, and where mammals move, even while accounting for seasonality (e.g, changes in photoperiod, hormones, and resources). Disentangling temperature from other drivers is important for predicting how changes in climate will affect mammalian movement when other drivers remain approximately constant. However, predicting mammals' responses to climate change becomes increasingly complex as habitats warm and animals are exposed to increasingly frequent conditions that are both extreme and novel. Predicting mammals' responses at the data-scarce fringes of the surface plots in Figs. \@ref(fig:dist) and \@ref(fig:hrsfs) comes with substantial uncertainty, particularly given that the responses are nonlinear. At unusually warm temperatures, mammals (and other homeotherms) overheat [@alston_temperature_2020; @dyer_travel_2023], so their movement behaviour and that of their competitors, predators, and/or prey can often be substantially different from the behaviour at typical temperatures. As extreme temperature events become more frequent and prolonged [@yao_emergence_2022; @ipcc_climate_2023], mammals will be increasingly forced towards atypical behaviours that may alter community structures and behaviours, both during such events and afterwards [@zhang_increased_2023; @logares_black_2012; @anderson_black-swan_2017]. For instance, changes in climate and phenology impact the life history and behaviour of many hibernating mammals [@wells_life_2022], and hot weather can also affects mammals' sleep quality [@mortlock_sleep_2024] and likelihood to enter torpor [@fjelldal_nightly_2021]. Such changes may alter the frequency and intensity of human-wildlife conflict, especially with the addition of growing pressures from human development and presence [@sih_evolution_2011; @johnson_human_2018; @weststrate_how_2024]. At the same time, warmer winters may reduce mammals' energetic expenditure [@berger_climate_2018; @schmidt_interplay_2020], increase ease of movement as snow cover and depth decrease [@leclerc_determinants_2021; @melin_effect_2023], increase their chances of finding food or being predated upon [@hou_cold_2020; @gilbert_behavioral_2017; @pedersen_quantifying_2021; @slatyer_ecological_2022; @sullender_defining_2023], and affect the timing and duration of migrations [@sawyer_identifying_2009; @leclerc_determinants_2021; @xu_plasticity_2021]. These changes will likely have complex consequences for population and ecosystem structures and dynamics as prey, predators, and competitors experience altered seasonal cycles and increasingly common climate "weirding" [@bunnell_global_2011].

Our ability to respond to current and future changes in climate is contingent on our ability to prepare for and predict change. However, predicting animal behaviour becomes increasingly complicated as the conditions animals are exposed to deviate from current, typical conditions, especially when responses are nonlinear and data are sparse. Consequently, we do not present our results as a definitive guide to how mammals in BC will respond to climate change. Instead, we hope they serve as a starting point to (1) demonstrate that mammals' movement rates and habitat selection depend on temperature and (2) how one can estimate mammals' changes in movement behaviour due to climate change. Additionally, communicating uncertainty in one's estimates is crucial in assessing risk probabilistically [@aven_implementing_2002; @ayre_bayesian_2012; @kluppelberg_bayesian_2014]. Recent events in global politics, including global conflicts [@mcnutt_scientists_2022; @tollefson_what_2022] and the rapid rise in generative machine learning [more commonly: "AI", see @roumeliotis_chatgpt_2023; @van_noorden_chatgpt_2023; @deepseek-ai_deepseek-r1_2025] and its prohibitive energy and water demands [@li_making_2023] and carbon footprint [@luccioni_estimating_2024], have placed doubt on the feasibility of reaching climate change goals that seemed achievable in the past decade [@united_nations_environment_programme_emissions_2024]. Mitigating climate change and its consequences will require extensive and widespread collaborative effort [@huang_achieving_2021]. Similarly, protecting a third of the world's ecosystems [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] appears less tangible as global political and environmental conditions change rapidly and unpredictably [@sih_evolution_2011; @simmons_tough_2021; @guenette_implications_2022; @sovacool_next_2023]. Achieving the "30 by 30" goal will require active partnership with local Peoples, especially Indigenous Peoples [@wong_towards_2020; @lamb_braiding_2023]. Understanding the consequences of climate change on mammals' movement behaviour and spatial distribution is a first step towards anticipating and proactively responding to future changes in human-induced rapid environmental change [@sih_evolution_2011; @williams_range_2018]. In the following sections, we discuss the implications of our results in more detail. We then expand on consequences for conservation during the 21^st^ century and considerations for future studies.

@@

Achieving the goals laid out by the "30 by 30" conservation initiative [Section H, Targets 2 and 3 of @convention_on_biological_diversity_154_2022] will require careful planning but often also rapid action. In the case of Canada, only 13.7% of its terrestrial areas and inland waters were protected at the end of 2023 [@environment_and_climate_change_canada_canadian_2024; @minister_of_environment_and_climate_change_act_2024]. Additionally, efficient and effective conservation will require collaboration with Indigenous and local communities to ensure Treaties are respected in the process [e.g., @falardeau_biophysical_2022; @lamb_braiding_2023; @wong_towards_2020; @turner_its_2009] while also actively preparing for future change [@desjardins_looking_2020; @hessami_indigenizing_2021]. The large range in predicted changes by 2100 (both across species and within species' current ranges) and high spatial heterogeneity in Fig. \@ref(fig:s-hrsfs) highlight the need for careful planning that incorporates not only reliable estimates of change but also pragmatic and communicable measures of uncertainty and variability around such estimates. The width of the prediction intervals in Fig. \@ref(fig:t-dist) suggest that spatial and intra-individual variation should be accounted for when deciding the location and size of protected areas. Accurate estimates of the effects of changes in temperature on mammals' movement require a holistic view of the direct effects of temperature on mammals' movement directly as well as its effects on other drivers of movement, such as forage and prey availability [@mezzini_how_2025], encounter rates [@hou_cold_2020; @martinez-garcia_how_2020], population dynamics [@smith_densitydependent_2023], competitive pressure [@torrez-herrera_monkeys_2020], and predation risk [@kohl_prey_2019]. This complexity results in the great variation among the responses of different species (and individuals), especially as populations a rapid cascade of change [@botero_evolutionary_2015] and animals face increasing pressures from human activity, including habitat fragmentation, habitat loss, and greater human presence in wild spaces [@sih_evolution_2011; @tucker_moving_2018; @sawyer_identifying_2009; @weststrate_how_2024; @rosenthal_impact_2022; @rice_conspicuous_2022]. As selection strength for current ranges changes (and likely decreases), some animals may disperse, relocate to a new habitat, or remain within the current range despite the reduced fitness and increased extinction risk [@logares_black_2012; @anderson_black-swan_2017; @duncan_predicting_2012].
